---
title: "BioHEART and Chimp merging"
author: "Andy Tran"
date: "University of Sydney | `r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    fig_caption: yes
    number_sections: yes
    self_contained: yes
    theme: flatly
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Data processing {.tabset}

## Load packages
```{r}
suppressPackageStartupMessages({
  library(plotly)
  library(tidyverse)
  # library(scMerge)
  library(ruv)
  library(impute)
  library(pheatmap)
  library(table1)
  library(ggpubr)
  })
```

## Load data and clean names

```{r}
load("data/bioheart/5_BH_Cleaned.RData")
bioheart_areas <- BH_MI$MetricData$Area %>% data.frame() %>% janitor::clean_names() %>% t %>% impute.knn()
bioheart_areas <- bioheart_areas$data %>% t %>% data.frame

bioheart_IS <- BH_MI$MetaboliteData 
bioheart_IS$IS_ind <- sapply(bioheart_IS$Final_ISTD_Name, function(x) which(bioheart_IS$Final_Metabolite_Name == x))  
dim(bioheart_areas)

chimp_areas <- readxl::read_excel("data/chimp/2023_03_29 - MP22_063_raw areas_mapped[1].xlsx") %>% data.frame()
chimp_IS <- readxl::read_excel("data/chimp/1_description/2023_tables for 6495 methods section.xlsx", skip = 1)
dim(chimp_IS)

rownames(chimp_areas) <- chimp_areas$Sample.ID
chimp_areas <- chimp_areas[,-(1:2)]
chimp_areas <- chimp_areas %>% janitor::clean_names()
dim(chimp_areas)
```

```{r}
lipid_name_mapping <- readxl::read_excel("data/2024_04_15 - Aligned BioHeart Nomenclature.xlsx")
lipid_name_mapping$BioHeart %>% na.omit() %>% length
lipid_name_mapping$Chimp %>% na.omit() %>% length
lipid_name_mapping$Common %>% na.omit() %>% length
```

```{r}
lipid_name_mapping <- lipid_name_mapping %>% 
  mutate(BioHeart_clean = janitor::make_clean_names(BioHeart),
         Chimp_clean = janitor::make_clean_names(Chimp),
         Common_clean = janitor::make_clean_names(Common)) 

setdiff(colnames(bioheart_areas), lipid_name_mapping$BioHeart_clean)
setdiff(lipid_name_mapping$BioHeart_clean, colnames(bioheart_areas))

setdiff(colnames(chimp_areas), lipid_name_mapping$Chimp_clean)
setdiff(lipid_name_mapping$Chimp_clean, colnames(chimp_areas))

comparison <- lipid_name_mapping %>% select(BioHeart, Chimp) %>% na.omit() 
sum(comparison$BioHeart == comparison$Chimp)
```

```{r}
colnames(bioheart_areas)[which(colnames(bioheart_areas) == "sulfatide_d18_1_16_0_oh")] <- "s_hex_cer_d18_1_16_0_oh"
colnames(bioheart_areas)[which(colnames(bioheart_areas) == "sulfatide_d18_1_16_0")] <- "s_hex_cer_d18_1_16_0"
colnames(bioheart_areas)[which(colnames(bioheart_areas) == "sulfatide_d18_1_24_0_oh")] <- "s_hex_cer_d18_1_24_0_oh"
colnames(bioheart_areas)[which(colnames(bioheart_areas) == "sulfatide_d18_1_24_0")] <- "s_hex_cer_d18_1_24_0"
colnames(bioheart_areas)[which(colnames(bioheart_areas) == "sulfatide_d18_1_24_1_oh")] <- "s_hex_cer_d18_1_24_1_oh"
colnames(bioheart_areas)[which(colnames(bioheart_areas) == "sulfatide_d18_1_24_1")] <- "s_hex_cer_d18_1_24_1"



bioheart_IS$Final_Metabolite_Name[which(bioheart_IS$Final_Metabolite_Name == "Sulfatide (d18:1:/16:0(OH))")] <- "SHexCer(d18:1/16:0(OH))"
bioheart_IS$Final_Metabolite_Name[which(bioheart_IS$Final_Metabolite_Name == "Sulfatide (d18:1:/16:0)")] <- "SHexCer(d18:1/16:0"
bioheart_IS$Final_Metabolite_Name[which(bioheart_IS$Final_Metabolite_Name == "Sulfatide (d18:1:/24:0(OH))")] <- "SHexCer(d18:1/24:0(OH))"
bioheart_IS$Final_Metabolite_Name[which(bioheart_IS$Final_Metabolite_Name == "Sulfatide (d18:1:/24:0)")] <- "SHexCer(d18:1/24:0"
bioheart_IS$Final_Metabolite_Name[which(bioheart_IS$Final_Metabolite_Name == "Sulfatide (d18:1:/24:1(OH))")] <- "SHexCer(d18:1/24:1(OH))"
bioheart_IS$Final_Metabolite_Name[which(bioheart_IS$Final_Metabolite_Name == "Sulfatide (d18:1:/24:1)")] <- "SHexCer(d18:1/24:1)"

```


```{r}
length(intersect(colnames(bioheart_areas), colnames(chimp_areas)))

update_ind <- which(colnames(bioheart_areas) %in% lipid_name_mapping$BioHeart_clean)
colnames(bioheart_areas)[update_ind] <- plyr::mapvalues(colnames(bioheart_areas)[update_ind],
                                                        from = lipid_name_mapping$BioHeart_clean,
                                                        to = lipid_name_mapping$Common_clean,
                                                        warn_missing = FALSE)


update_ind <- which(colnames(chimp_areas) %in% lipid_name_mapping$Chimp_clean)
colnames(chimp_areas)[update_ind] <- plyr::mapvalues(colnames(chimp_areas)[update_ind],
                                                        from = lipid_name_mapping$Chimp_clean,
                                                        to = lipid_name_mapping$Common_clean,
                                                        warn_missing = FALSE)

length(intersect(colnames(bioheart_areas), colnames(chimp_areas)))
```


## Bioheart tube labels and clinical data

```{r}
labels <- read.csv("data/bioheart/tube_label.csv")
clinical_data_full <- readxl::read_excel("data/bioheart/BioCT-DL_20220209.xlsx")

tubes_present <- unname(sapply(rownames(bioheart_areas), function(x) strsplit(x, "_")[[1]][5]))
tubes_present <- sub(" ", "", tubes_present)
labels$tube_id_clean <- sub(" ", "", labels$tube_id)

present_samples <- plyr::mapvalues(tubes_present, 
                                   from = labels$tube_id_clean, 
                                   to = labels$sample.ID, 
                                   warn_missing = FALSE)
  
clinical_data <- data.frame(record_id = as.numeric(present_samples[BH_MI$SampleData$SampleTypeLabel == "BIO"]))
clinical_data <- mutate(clinical_data, tube_id_clean = plyr::mapvalues(record_id, 
                                                                       from = labels$sample.ID, 
                                                                       to = labels$tube_id_clean, 
                                                                       warn_missing = FALSE))
clinical_data <- left_join(clinical_data, clinical_data_full, by = "record_id")
```



# Calculate Concentrations using IS {.tabset}

## Checking IS names

```{r}
IS_present_chimp <- data.frame(ind = which(str_detect(colnames(chimp_areas), "is$"))) %>%
  mutate(name = colnames(chimp_areas)[ind],
         class = sub("\\_.*", "", name)) 

chimp_IS <- chimp_IS %>% mutate(name_clean = janitor::make_clean_names(chimp_IS$`Internal Standard`),
                                name_clean = paste0(name_clean, "_is"))
chimp_IS <- chimp_IS %>% 
  mutate(name_clean = plyr::mapvalues(chimp_IS$name_clean,
                                      from = c("ac_16_0_d3_is", 
                                               "dh_cer_d18_0_8_0_is", 
                                               "ffa_18_1_d9_is", 
                                               "s1p_18_1_d7_is",
                                               "hex_cer_d18_1_d7_15_0_is",
                                               "tg_15_0_18_1_d7_15_0_2_is",
                                               "hex2cer_d18_1_d7_15_0_is",
                                               "sm_d18_1_d9_15_0_is"),
                                      to = c("acyl_carnitine_16_0_d3_is", 
                                             "dh_cer_8_0_is", 
                                             "fa_18_1_d9_is", 
                                             "s1p_d18_1_d7_is",
                                             "hex_cer_d18_1_15_0_d7_is",
                                             "tg_48_1_nl_18_1_d7_is",
                                             "hex2cer_d18_1_15_0_d7_is",
                                             "sm_d18_1_15_0_d9_is"))) %>%
  mutate(class = sub("\\_.*", "", name_clean))
```

## Create lipid to IS mapping

```{r}
ISTD_matchup <- readxl::read_excel("data/ISTD_matchup.xlsx") %>% 
  data.frame() %>%
  mutate(analyte_clean = janitor::make_clean_names(analyte),
         ISTD_clean = sapply(ISTD, function(x) janitor::make_clean_names(x)))

#setdiff(colnames(chimp_areas), ISTD_matchup$analyte_clean)

IS_mapping <- data.frame(analyte = colnames(chimp_areas)) %>%
  mutate(IS = plyr::mapvalues(analyte, from = ISTD_matchup$analyte_clean, to = ISTD_matchup$ISTD_clean, warn_missing = F),
         IS_ind = plyr::mapvalues(IS, from = IS_present_chimp$name, to = IS_present_chimp$ind, warn_missing = F),
         IS_ind = as.numeric(IS_ind),
         IS_conc = plyr::mapvalues(IS, from = chimp_IS$name_clean, to = chimp_IS$`Internal Standard (pmol)/sample`, warn_missing = F),
         IS_conc = as.numeric(sub(",", "", IS_conc)),
         class = sub("\\_.*", "", IS))
  
lipid_classes <- unique(IS_mapping$class)

## remove lipids that don't match to an IS or aren't found in the Baker IS list
missing_inds <- is.na(IS_mapping$IS_ind) | is.na(IS_mapping$IS_conc)
chimp_areas <- chimp_areas[,!missing_inds]
IS_mapping <- IS_mapping[!missing_inds,]

dim(chimp_areas)
dim(IS_mapping)
```

```{r}
IS_present_bioheart <- data.frame(ind = which(str_detect(colnames(bioheart_areas), "is$"))) %>%
  mutate(name = colnames(bioheart_areas)[ind],
         class = sub("\\_.*", "", name)) 

intersect(IS_present_chimp$name, IS_present_bioheart$name)
setdiff(IS_present_chimp$name, IS_present_bioheart$name)
setdiff(IS_present_bioheart$name, IS_present_chimp$name)

bioheart_IS <- bioheart_IS %>% 
  mutate(analyte = janitor::make_clean_names(Final_Metabolite_Name),
         ISTD_clean = sapply(Final_ISTD_Name, function(x) janitor::make_clean_names(x)),
         class = sub("\\_.*", "", ISTD_clean))
```


## Calculate concentrations

```{r}
chimp_matrix <- as.matrix(chimp_areas)
chimp_conc <- matrix(nrow = nrow(chimp_matrix), ncol = ncol(chimp_matrix))
for(i in 1:nrow(chimp_matrix)){
  chimp_conc[i,] <- (chimp_matrix[i,]/chimp_matrix[i,IS_mapping$IS_ind]) * IS_mapping$IS_conc *100
}
dimnames(chimp_conc) <- dimnames(chimp_areas)
chimp_conc <- data.frame(chimp_conc)
dim(chimp_conc)
```

```{r}
bioheart_matrix <- as.matrix(bioheart_areas)
bioheart_conc <- matrix(nrow = nrow(bioheart_matrix), ncol = ncol(bioheart_matrix))
for(i in 1:nrow(bioheart_matrix)){
  bioheart_conc[i,] <- (bioheart_matrix[i,]/bioheart_matrix[i,bioheart_IS$IS_ind]) * bioheart_IS$Ref_Concentration * bioheart_IS$Response_Factor *100
}
dimnames(bioheart_conc) <- dimnames(bioheart_areas)
bioheart_conc <- data.frame(bioheart_conc)
dim(bioheart_conc)
```


```{r}
bioheart_conc_long <- pivot_longer(bioheart_conc, cols = everything(), names_to = "analyte", values_to = "New_Value")

old_conc_long <- BH_MI$MetricData$Conc %>% 
  data.frame() %>% 
  janitor::clean_names() %>%
  pivot_longer(cols = everything(), names_to = "analyte", values_to = "Old_Value")

identical(bioheart_conc_long$analyte, old_conc_long$analyte)

plot_df <- bioheart_conc_long
plot_df$Old_Value <- old_conc_long$Old_Value
plot_df <- plot_df %>% 
  group_by(analyte) %>%
  summarise(New_Value = mean(New_Value), Old_Value = mean(Old_Value)) %>%
  mutate(class = plyr::mapvalues(analyte, from = bioheart_IS$analyte, to = bioheart_IS$Metabolite_Class, warn_missing = F))

p <- plot_df %>% 
  ggplot(aes(x = log(Old_Value),
             y = log(New_Value), 
             color = class,
             text = paste("Analyte:", analyte))) + 
  geom_point() +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed") +
  labs(title="Observed analyte means between original and calculated concentration calculations",
       x ="Original Log Values", 
       y = "Calculated Log values")
ggplotly(p)
```


# Set up {.tabset}

## Subset to common lipids

```{r}
## filter to common lipids
common_lipids_withIS <- sort(Reduce(intersect, list(colnames(chimp_conc),colnames(bioheart_areas))))
length(common_lipids_withIS)

chimp_areas_common_withIS <- chimp_areas[,common_lipids_withIS]
bioheart_areas_common_withIS <- bioheart_areas[,common_lipids_withIS]


common_lipids <- common_lipids_withIS[!str_detect(common_lipids_withIS, "is$")]  ## remove IS lipids
length(common_lipids)

chimp_conc_common <- chimp_conc[,common_lipids]
bioheart_conc_common <- bioheart_conc[,common_lipids]
chimp_areas_common <- chimp_areas[,common_lipids]
bioheart_areas_common <- bioheart_areas[,common_lipids] 
```

```{r}
setdiff(lipid_name_mapping$Common_clean, common_lipids)
```


Create mapping 

```{r}
class_mapping <- bioheart_IS %>% select(analyte, Metabolite_Class) %>% filter(analyte %in% common_lipids) %>% arrange(analyte)

class_mapping <- class_mapping[order(class_mapping$analyte),]

identical(colnames(bioheart_areas_common), colnames(chimp_areas_common))
identical(colnames(bioheart_conc_common), colnames(chimp_conc_common))
identical(class_mapping$analyte, colnames(bioheart_conc_common))
```


## Define functions

Check distribution function

```{r}
check_distribution <- function(bioheart_data, chimp_data){
  bioheart_long <- bioheart_data %>% 
  mutate(sampleID = rownames(bioheart_data)) %>%
  pivot_longer(cols = -"sampleID", names_to = "species", values_to  = "Value") %>%
  mutate(cohort = "bioheart")

chimp_long <- chimp_data %>% 
  mutate(sampleID = rownames(chimp_data)) %>%
  pivot_longer(cols = -"sampleID", names_to = "species", values_to  = "Value") %>%
  mutate(cohort = "chimp")

combined_long <- rbind(bioheart_long, chimp_long) %>% 
  mutate(cohort = factor(cohort, levels = c("bioheart", "chimp")),
         class = plyr::mapvalues(species, from = class_mapping$analyte, to = class_mapping$Metabolite_Class, warn_missing = F))

lipid_totals_order <- combined_long %>% group_by(class) %>% summarise(sums = sum(Value)) %>% arrange(-sums) %>% pull(class)

combined_long %>%  
  mutate(class = factor(class, levels = lipid_totals_order)) %>%
  group_by(class, cohort, sampleID) %>%
  summarise(Value = sum(Value)) %>%
  ggplot(aes(x = class, y = log2(Value), colour = cohort)) + 
    geom_boxplot() + 
    theme_bw() +  
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
    ggtitle("Totals by class")
}
```

PCA visualisation

```{r}
pca_visualisation <- function(bioheart_data, chimp_data){
  
combined_bc <- log2(rbind(bioheart_data, chimp_data))
batch = c(rep("bioheart", nrow(bioheart_data)), 
          rep("chimp", nrow(chimp_data)))

pca_combined = prcomp(combined_bc)
df_pca <- data.frame(pca_combined$x)
df_pca$cohort <- batch

prop_var <- 100*(round(data.frame(summary(pca_combined)$importance)[2,], 4))

ggplot(df_pca, aes(x = PC1, y = PC2, colour = cohort)) +
  geom_point() +
  xlab(paste0("PC1 (", prop_var$PC1, "%)")) + 
  ylab(paste0("PC2 (", prop_var$PC2, "%)")) +
  ggtitle("PCA visualisation") +
  theme_bw() 
}
```

PCA visualisation with variables

```{r}
pca_visualisation_detailed <- function(bioheart_data, chimp_data){
  
combined_bc <- log2(rbind(bioheart_data, chimp_data))
batch = c(rep("bioheart", nrow(bioheart_data)), 
          rep("chimp", nrow(chimp_data)))

pca_combined = prcomp(combined_bc)
df_pca <- data.frame(pca_combined$x)
df_pca$cohort <- batch

tubes_present <- unname(sapply(rownames(bioheart_data), function(x) strsplit(x, "_")[[1]][5]))
tubes_present <- sub(" ", "", tubes_present)
tubes_present <- c(tubes_present, rep(NA, nrow(chimp_data)))
df_pca$tube_id_clean <- tubes_present
df_pca <- left_join(df_pca, clinical_data, by = "tube_id_clean")

prop_var <- 100*(round(data.frame(summary(pca_combined)$importance)[2,], 4))

p1 <- ggplot(df_pca, aes(x = PC1, y = PC2, colour = cohort)) +
  geom_point() +
  xlab(paste0("PC1 (", prop_var$PC1, "%)")) + 
  ylab(paste0("PC2 (", prop_var$PC2, "%)")) +
  ggtitle("PCA visualisation") +
  theme_bw() 

p2 <- ggplot(df_pca, aes(x = PC1, y = PC2, colour = age)) +
  geom_point() + 
  scale_colour_viridis_c() +
  xlab(paste0("PC1 (", prop_var$PC1, "%)")) + 
  ylab(paste0("PC2 (", prop_var$PC2, "%)")) +
  ggtitle("PCA visualisation") +
  theme_bw() 

p3 <- ggplot(df_pca, aes(x = PC1, y = PC2, colour = factor(gender))) +
  geom_point() + 
  xlab(paste0("PC1 (", prop_var$PC1, "%)")) + 
  ylab(paste0("PC2 (", prop_var$PC2, "%)")) +
  ggtitle("PCA visualisation") +
  theme_bw() 

p4 <- ggplot(df_pca, aes(x = PC1, y = PC2, colour = cacs>0)) +
  geom_point() + 
  xlab(paste0("PC1 (", prop_var$PC1, "%)")) + 
  ylab(paste0("PC2 (", prop_var$PC2, "%)")) +
  ggtitle("PCA visualisation") +
  theme_bw() 

p5 <- ggplot(df_pca, aes(x = PC1, y = PC2, colour = as.numeric(gensini)>0)) +
  geom_point() + 
  xlab(paste0("PC1 (", prop_var$PC1, "%)")) + 
  ylab(paste0("PC2 (", prop_var$PC2, "%)")) +
  ggtitle("PCA visualisation") +
  theme_bw() 

print(p1)
print(p2)
print(p3)
print(p4)
print(p5)
}

```

Plot correlations

```{r}
plot_correlations <- function(bioheart_data, chimp_data){
  cor_conc <- cor(log2(t(bioheart_data)), log2(t(chimp_data)))
  pheatmap(cor_conc, show_rownames = F, main = "Correlation between bioheart and chimp samples")
}
```

Plot differences

```{r}
plot_differences <- function(bioheart_data, chimp_data){
  plot_df <- data.frame(means_bioheart = log2(colMeans(bioheart_data)),
                        means_chimp = log2(colMeans(chimp_data)),
                        analyte = colnames(chimp_data))
  plot_df <- plot_df %>% 
    mutate(class = plyr::mapvalues(analyte, from = class_mapping$analyte, to = class_mapping$Metabolite_Class, warn_missing = F))
  
  p <- plot_df %>% 
    ggplot(aes(x = means_chimp,
               y = means_bioheart, 
               color = class,
               text = paste("Analyte:", analyte))) + 
    geom_point() +
    geom_abline(intercept = 0, slope = 1, linetype = "dashed") +
    labs(title="Observed analyte means between bioheart and chimp cohorts",
         x ="Chimp Means", 
         y = "Bioheart Means")
  ggplotly(p)
}
```


Extract NIST and BIO samples

```{r}
extract_NIST_bioheart <- function(bioheart_data){bioheart_data[BH_MI$SampleData$SampleTypeLabel == "NIST1950",]}
extract_BIO_bioheart <- function(bioheart_data){bioheart_data[BH_MI$SampleData$SampleTypeLabel == "BIO",]}
extract_NIST_chimp <- function(chimp_data){chimp_data[str_which(rownames(chimp_data),"NIST"),]}
extract_BIO_chimp <- function(chimp_data){chimp_data[str_which(rownames(chimp_data),"TZ"),]}
```






# Visualise Raw data {.tabset}

## NIST areas
```{r}
bioheart_areas_nist <- extract_NIST_bioheart(bioheart_areas_common)
chimp_areas_nist <- extract_NIST_chimp(chimp_areas_common)
check_distribution(bioheart_areas_nist, chimp_areas_nist)
pca_visualisation(bioheart_areas_nist, chimp_areas_nist)
plot_correlations(bioheart_areas_nist, chimp_areas_nist)
plot_differences(bioheart_areas_nist, chimp_areas_nist)
```

## BIO areas
```{r}
bioheart_areas_BIO <- extract_BIO_bioheart(bioheart_areas_common)
chimp_areas_BIO <- extract_BIO_chimp(chimp_areas_common)  
check_distribution(bioheart_areas_BIO, chimp_areas_BIO)
pca_visualisation_detailed(bioheart_areas_BIO, chimp_areas_BIO)
plot_correlations(bioheart_areas_BIO, chimp_areas_BIO)
plot_differences(bioheart_areas_BIO, chimp_areas_BIO)
```


## NIST concentrations
```{r}
bioheart_conc_nist <- extract_NIST_bioheart(bioheart_conc_common)
chimp_conc_nist <- extract_NIST_chimp(chimp_conc_common)
check_distribution(bioheart_conc_nist, chimp_conc_nist)
pca_visualisation(bioheart_conc_nist, chimp_conc_nist)
plot_correlations(bioheart_conc_nist, chimp_conc_nist)
plot_differences(bioheart_conc_nist, chimp_conc_nist)
```

## BIO concentrations
```{r}
bioheart_conc_BIO <- extract_BIO_bioheart(bioheart_conc_common)
chimp_conc_BIO <- extract_BIO_chimp(chimp_conc_common)  
check_distribution(bioheart_conc_BIO, chimp_conc_BIO)
pca_visualisation_detailed(bioheart_conc_BIO, chimp_conc_BIO)
plot_correlations(bioheart_conc_BIO, chimp_conc_BIO)
plot_differences(bioheart_conc_BIO, chimp_conc_BIO)
```


# Checking QC of Raw data sets {.tabset}

## BioHEART {.tabset}

### BioHEART QC funtions

```{r}
identical(rownames(BH_MI$SampleData), rownames(bioheart_areas_common))

table(BH_MI$SampleData$SampleTypeLabel)
table(BH_MI$SampleData$Batch)
```


```{r}
bioheart_qc_IS <- function(bioheart_data_withIS, IS_name){
  bioheart_data_withIS %>% 
  mutate(sampleID = rownames(bioheart_data_withIS),
         order = BH_MI$SampleData$InjectionOrder,
         sampleType = BH_MI$SampleData$SampleTypeLabel) %>%
  pivot_longer(cols = -c("sampleID", "order", "sampleType"), names_to = "species", values_to  = "Value") %>% 
  filter(species == IS_name) %>%
  ggplot(aes(x = order, y = log2(Value), colour = sampleType)) + 
    geom_line() + 
    theme_bw() +  
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
    ggtitle(paste(IS_name, "for sample types"))
}

bioheart_qc_classTotals <- function(bioheart_data, class_name){
  bioheart_data %>% 
  mutate(sampleID = rownames(bioheart_data),
         order = BH_MI$SampleData$InjectionOrder,
         sampleType = BH_MI$SampleData$SampleTypeLabel) %>%
  pivot_longer(cols = -c("sampleID", "order", "sampleType"), names_to = "species", values_to  = "Value") %>% 
  mutate(class = plyr::mapvalues(species, from = class_mapping$analyte, to = class_mapping$Metabolite_Class, warn_missing = F)) %>%  
  group_by(class, sampleID, order, sampleType) %>%
  summarise(Value = sum(Value)) %>%
  filter(class == class_name) %>%
  ggplot(aes(x = order, y = log2(Value), colour = sampleType)) + 
    geom_line() + 
    theme_bw() +  
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
    ggtitle(paste(class_name, "class totals over time for sample types"))
}

bioheart_qc_sampleType <- function(bioheart_data, sampleType_name){
  bioheart_data %>% 
  mutate(sampleID = rownames(bioheart_data),
         order = BH_MI$SampleData$InjectionOrder,
         sampleType = BH_MI$SampleData$SampleTypeLabel) %>%
  pivot_longer(cols = -c("sampleID", "order", "sampleType"), names_to = "species", values_to  = "Value") %>% 
  mutate(class = plyr::mapvalues(species, from = class_mapping$analyte, to = class_mapping$Metabolite_Class, warn_missing = F)) %>%  
  group_by(class, sampleID, order, sampleType) %>%
  summarise(Value = sum(Value)) %>%
  filter(sampleType == sampleType_name) %>%
  ggplot(aes(x = order, y = log2(Value), colour = class)) + 
    geom_line() + 
    theme_bw() +  
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
    ggtitle(paste("Class totals over time for", sampleType_name))
}
```

### BioHEART IS QC

```{r}
bioheart_qc_IS(bioheart_areas_common_withIS, "acyl_carnitine_16_0_d3_is")
bioheart_qc_IS(bioheart_areas_common_withIS, "cer_d18_1_d7_18_0_is")
bioheart_qc_IS(bioheart_areas_common_withIS, "sph_d17_1_is")
bioheart_qc_IS(bioheart_areas_common_withIS, "fa_18_1_d9_is")
bioheart_qc_IS(bioheart_areas_common_withIS, "lpi_13_0_is")
```

### BioHEART Class totals over time

```{r}
bioheart_qc_classTotals(bioheart_areas_common, "PC")
bioheart_qc_classTotals(bioheart_areas_common, "TG [NL]")
bioheart_qc_classTotals(bioheart_areas_common, "Cer(m)")
bioheart_qc_classTotals(bioheart_areas_common, "COH")
bioheart_qc_classTotals(bioheart_areas_common, "Hex3Cer")
bioheart_qc_classTotals(bioheart_areas_common, "GM1")
```


### BioHEART sample types over time

```{r}
bioheart_qc_sampleType(bioheart_areas_common, "BIO")
bioheart_qc_sampleType(bioheart_areas_common, "PQC")
bioheart_qc_sampleType(bioheart_areas_common, "TQC")
bioheart_qc_sampleType(bioheart_areas_common, "NIST1950")
bioheart_qc_sampleType(bioheart_areas_common, "BLANK+")
```

## Chimp {.tabset}

### Chimp QC funtions

```{r}
chimp_qc_IS <- function(chimp_data_withIS, IS_name){
  chimp_data_withIS %>% 
  mutate(sampleID = rownames(chimp_data_withIS),
         sampleType = sub("( |-).*$", "", rownames(chimp_data_withIS))) %>%
  pivot_longer(cols = -c("sampleID", "sampleType"), names_to = "species", values_to  = "Value") %>%
  filter(species == IS_name) %>%
  ggplot(aes(x = sampleType, y = log2(Value), colour = sampleType)) + 
    geom_boxplot() + 
    theme_bw() +  
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
    ggtitle(paste(IS_name, "for sample types"))
}

chimp_qc_classTotals <- function(chimp_data, class_name){
  chimp_data %>% 
  mutate(sampleID = rownames(chimp_data),
         sampleType = sub("( |-).*$", "", rownames(chimp_data))) %>%
  pivot_longer(cols = -c("sampleID", "sampleType"), names_to = "species", values_to  = "Value") %>% 
  mutate(class = plyr::mapvalues(species, from = class_mapping$analyte, to = class_mapping$Metabolite_Class, warn_missing = F)) %>%  
  group_by(class, sampleID, sampleType) %>%
  summarise(Value = sum(Value)) %>%
  filter(class == class_name) %>%
  ggplot(aes(x = sampleType, y = log2(Value), colour = sampleType)) + 
    geom_boxplot() + 
    theme_bw() +  
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
    ggtitle(paste(class_name, "class totals for sample types"))
}
```

### Chimp IS QC

```{r}
chimp_qc_IS(chimp_areas_common_withIS, "acyl_carnitine_16_0_d3_is")
chimp_qc_IS(chimp_areas_common_withIS, "cer_d18_1_d7_18_0_is")
chimp_qc_IS(chimp_areas_common_withIS, "sph_d17_1_is")
chimp_qc_IS(chimp_areas_common_withIS, "fa_18_1_d9_is")
chimp_qc_IS(chimp_areas_common_withIS, "lpi_13_0_is")
```

### Chimp Class totals over sample type

```{r}
chimp_qc_classTotals(chimp_areas_common, "PC")
chimp_qc_classTotals(chimp_areas_common, "TG [NL]")
chimp_qc_classTotals(chimp_areas_common, "Cer(m)")
chimp_qc_classTotals(chimp_areas_common, "COH")
chimp_qc_classTotals(chimp_areas_common, "Hex3Cer")
chimp_qc_classTotals(chimp_areas_common, "GM1")
```



# Merge within data sets {.tabset}

```{r}
source("data/RUVIII_SPB.R")
```

## BioHEART all at once {.tabset}

```{r}
Y <- log2(bioheart_areas_common_withIS+1)
M <- 1:nrow(Y)

NIST_ind_bioheart <- which(BH_MI$SampleData$SampleTypeLabel == "NIST1950")
M[NIST_ind_bioheart] <- "NIST"

TQC_ind_bioheart <- which(BH_MI$SampleData$SampleTypeLabel == "TQC")
M[TQC_ind_bioheart] <- "TQC_bioheart"

PQC_ind_bioheart <- which(BH_MI$SampleData$SampleTypeLabel == "PQC")
M[PQC_ind_bioheart] <- "PQC_bioheart"

ctl <- str_detect(colnames(Y), "is$")
# ctl <- 1:ncol(Y)

new_Y <- RUVIII_SPB(Y, M, ctl, k = 3)
new_Y <- 2^new_Y
bioheart_areas_norm_withIS <- new_Y %>% data.frame()
bioheart_areas_norm <- bioheart_areas_norm_withIS[,common_lipids]
```

### BioHEART IS QC

```{r}
bioheart_qc_IS(bioheart_areas_norm_withIS, "acyl_carnitine_16_0_d3_is")
bioheart_qc_IS(bioheart_areas_norm_withIS, "cer_d18_1_d7_18_0_is")
bioheart_qc_IS(bioheart_areas_norm_withIS, "sph_d17_1_is")
bioheart_qc_IS(bioheart_areas_norm_withIS, "fa_18_1_d9_is")
bioheart_qc_IS(bioheart_areas_norm_withIS, "lpi_13_0_is")
```

### BioHEART Class totals over time

```{r}
bioheart_qc_classTotals(bioheart_areas_norm, "PC")
bioheart_qc_classTotals(bioheart_areas_norm, "TG [NL]")
bioheart_qc_classTotals(bioheart_areas_norm, "Cer(m)")
bioheart_qc_classTotals(bioheart_areas_norm, "COH")
bioheart_qc_classTotals(bioheart_areas_norm, "Hex3Cer")
bioheart_qc_classTotals(bioheart_areas_norm, "GM1")
```


### BioHEART sample types over time

```{r}
bioheart_qc_sampleType(bioheart_areas_norm, "BIO")
bioheart_qc_sampleType(bioheart_areas_norm, "PQC")
bioheart_qc_sampleType(bioheart_areas_norm, "TQC")
bioheart_qc_sampleType(bioheart_areas_norm, "NIST1950")
bioheart_qc_sampleType(bioheart_areas_norm, "BLANK+")
```

## Chimp {.tabset}

```{r}
Y <- log2(chimp_areas_common_withIS+1)
M <- 1:nrow(Y)

NIST_ind_chimp <- str_which(rownames(chimp_areas_common_withIS),"NIST")
M[NIST_ind_chimp] <- "NIST"

TQC_ind_chimp <- str_which(rownames(chimp_areas_common_withIS),"TQC")
M[TQC_ind_chimp] <- "TQC_chimp"

PQC_ind_chimp <- str_which(rownames(chimp_areas_common_withIS),"PQC")
M[PQC_ind_chimp] <- "PQC_chimp"

ctl <- str_detect(colnames(Y), "is$")
# ctl <- 1:ncol(Y)

new_Y <- RUVIII_SPB(Y, M, ctl, k = 3)
new_Y <- 2^new_Y
chimp_areas_norm_withIS <- new_Y %>% data.frame()
chimp_areas_norm <- chimp_areas_norm_withIS[,common_lipids]
```

### chimp IS QC

```{r}
chimp_qc_IS(chimp_areas_norm_withIS, "acyl_carnitine_16_0_d3_is")
chimp_qc_IS(chimp_areas_norm_withIS, "cer_d18_1_d7_18_0_is")
chimp_qc_IS(chimp_areas_norm_withIS, "sph_d17_1_is")
chimp_qc_IS(chimp_areas_norm_withIS, "fa_18_1_d9_is")
chimp_qc_IS(chimp_areas_norm_withIS, "lpi_13_0_is")
```

### chimp Class totals across sample type

```{r}
chimp_qc_classTotals(chimp_areas_norm, "PC")
chimp_qc_classTotals(chimp_areas_norm, "TG [NL]")
chimp_qc_classTotals(chimp_areas_norm, "Cer(m)")
chimp_qc_classTotals(chimp_areas_norm, "COH")
chimp_qc_classTotals(chimp_areas_norm, "Hex3Cer")
chimp_qc_classTotals(chimp_areas_norm, "GM1")
```



# Merging data {.tabset}

Merge on areas using RUVSPB with k = 3, with NIST, replicates, common IS as negative controls 

```{r}
Y <- log2(rbind(bioheart_areas_norm_withIS, chimp_areas_norm_withIS)+1)
M <- 1:nrow(Y)

NIST_ind_bioheart <- which(BH_MI$SampleData$SampleTypeLabel == "NIST1950")
NIST_ind_chimp <- str_which(rownames(chimp_areas_norm_withIS),"NIST")
NIST_inds <- c(NIST_ind_bioheart, NIST_ind_chimp + nrow(bioheart_areas_norm_withIS))
M[NIST_inds] <- "NIST"

ctl <- str_detect(colnames(Y), "is$")
# ctl <- 1:ncol(Y)

new_Y <- RUVIII_SPB(Y, M, ctl, k = 3)
new_Y <- 2^new_Y

new_bioheart_withIS <- new_Y[1:nrow(bioheart_areas_common_withIS),] %>% data.frame()
new_bioheart <- new_bioheart_withIS[,common_lipids]

new_chimp_withIS <- new_Y[-(1:nrow(bioheart_areas_norm_withIS)),] %>% data.frame()
new_chimp <- new_chimp_withIS[,common_lipids]
```

## BioHEART IS QC

```{r}
bioheart_qc_IS(new_bioheart_withIS, "acyl_carnitine_16_0_d3_is")
bioheart_qc_IS(new_bioheart_withIS, "cer_d18_1_d7_18_0_is")
bioheart_qc_IS(new_bioheart_withIS, "sph_d17_1_is")
bioheart_qc_IS(new_bioheart_withIS, "fa_18_1_d9_is")
bioheart_qc_IS(new_bioheart_withIS, "lpi_13_0_is")
```

## BioHEART Class totals over time

```{r}
bioheart_qc_classTotals(new_bioheart, "PC")
bioheart_qc_classTotals(new_bioheart, "TG [NL]")
bioheart_qc_classTotals(new_bioheart, "Cer(m)")
bioheart_qc_classTotals(new_bioheart, "COH")
bioheart_qc_classTotals(new_bioheart, "Hex3Cer")
bioheart_qc_classTotals(new_bioheart, "GM1")
```


## BioHEART sample types over time

```{r}
bioheart_qc_sampleType(new_bioheart, "BIO")
bioheart_qc_sampleType(new_bioheart, "PQC")
bioheart_qc_sampleType(new_bioheart, "TQC")
bioheart_qc_sampleType(new_bioheart, "NIST1950")
bioheart_qc_sampleType(new_bioheart, "BLANK+")
```

## chimp IS QC

```{r}
chimp_qc_IS(new_chimp_withIS, "acyl_carnitine_16_0_d3_is")
chimp_qc_IS(new_chimp_withIS, "cer_d18_1_d7_18_0_is")
chimp_qc_IS(new_chimp_withIS, "sph_d17_1_is")
chimp_qc_IS(new_chimp_withIS, "fa_18_1_d9_is")
chimp_qc_IS(new_chimp_withIS, "lpi_13_0_is")
```

## chimp Class totals across sample type

```{r}
chimp_qc_classTotals(new_chimp, "PC")
chimp_qc_classTotals(new_chimp, "TG [NL]")
chimp_qc_classTotals(new_chimp, "Cer(m)")
chimp_qc_classTotals(new_chimp, "COH")
chimp_qc_classTotals(new_chimp, "Hex3Cer")
chimp_qc_classTotals(new_chimp, "GM1")
```

## check NIST

```{r}
bioheart_nist <- extract_NIST_bioheart(new_bioheart)
chimp_nist <- extract_NIST_chimp(new_chimp)
check_distribution(bioheart_nist, chimp_nist)
pca_visualisation(bioheart_nist, chimp_nist)
plot_correlations(bioheart_nist, chimp_nist)
plot_differences(bioheart_nist, chimp_nist)
```

## check BIO

```{r}
bioheart_BIO <- extract_BIO_bioheart(new_bioheart)
chimp_BIO <- extract_BIO_chimp(new_chimp)
check_distribution(bioheart_BIO, chimp_BIO)
pca_visualisation_detailed(bioheart_BIO, chimp_BIO)
plot_correlations(bioheart_BIO, chimp_BIO)
plot_differences(bioheart_BIO, chimp_BIO)
```



## Check PCA with NIST

```{r}
combined_bc <- log2(rbind(bioheart_BIO, chimp_BIO, bioheart_nist, chimp_nist))
batch = c(rep("bioheart", nrow(bioheart_BIO)), 
          rep("chimp", nrow(chimp_BIO)),
          rep("NIST", nrow(bioheart_nist) + nrow(chimp_nist)))

pca_combined = prcomp(combined_bc)
df_pca <- data.frame(pca_combined$x)
df_pca$cohort <- batch

prop_var <- 100*(round(data.frame(summary(pca_combined)$importance)[2,], 4))

ggplot(df_pca, aes(x = PC1, y = PC2, colour = cohort)) +
  geom_point() +
  xlab(paste0("PC1 (", prop_var$PC1, "%)")) + 
  ylab(paste0("PC2 (", prop_var$PC2, "%)")) +
  ggtitle("PCA visualisation") +
  theme_bw() +
  scale_color_manual(values = c("firebrick1", "dodgerblue1", "darkgreen"))
```



## Check boxes

```{r}
combined_bc <- log2(rbind(bioheart_BIO, chimp_BIO))
batch = c(rep("bioheart", nrow(bioheart_BIO)), 
          rep("chimp", nrow(chimp_BIO)))

pca_combined = prcomp(combined_bc)
df_pca <- data.frame(pca_combined$x)
df_pca$cohort <- batch

tubes_present <- unname(sapply(rownames(bioheart_BIO), function(x) strsplit(x, "_")[[1]][5]))
tubes_present <- sub(" ", "", tubes_present)
tubes_present <- c(tubes_present, rep(NA, nrow(chimp_BIO)))
df_pca$tube_id_clean <- tubes_present
df_pca$box <- plyr::mapvalues(df_pca$tube_id_clean, from = labels$tube_id_clean, to = labels$Box, warn_missing = F)
df_pca <- left_join(df_pca, clinical_data, by = "tube_id_clean")

prop_var <- 100*(round(data.frame(summary(pca_combined)$importance)[2,], 4))

ggplot(df_pca, aes(x = PC1, y = PC2, colour = box)) +
  geom_point() +
  xlab(paste0("PC1 (", prop_var$PC1, "%)")) + 
  ylab(paste0("PC2 (", prop_var$PC2, "%)")) +
  ggtitle("PCA visualisation") +
  theme_bw() 
```


```{r, eval = FALSE}
saveRDS(bioheart_BIO, "data/merged/bioheart_BIO_is_20240421.rds")
saveRDS(chimp_BIO, "data/merged/chimp_BIO_is_20240421.rds")
saveRDS(class_mapping, "data/merged/class_mapping.rds")
saveRDS(lipid_name_mapping, "data/merged/lipid_name_mapping.rds")
```


# Checking against clinical factors {.tabset}

## set up

```{r}
identical(dimnames(bioheart_areas_BIO), dimnames(bioheart_BIO))
bioheart_removed_variation <- log2(bioheart_BIO) - log2(bioheart_areas_BIO)
bioheart_removed_variation[1:5,1:5]

bioheart_SampleData <- readRDS("data/merged/bioheart_SampleData.rds")
```

```{r}
areas_records <- plyr::mapvalues(rownames(bioheart_removed_variation),
                                          from = bioheart_SampleData$InjectionLabel,
                                          to = sub(" ", "", bioheart_SampleData$Other),
                                          warn_missing = FALSE)

areas_records <- plyr::mapvalues(areas_records,
                                          from = clinical_data$tube_id_clean,
                                          to = clinical_data$record_id,
                                          warn_missing = FALSE)

rownames(bioheart_removed_variation) <- areas_records
```

```{r}
key_clinical_vars <- clinical_data %>% select(record_id, age, gender, cacs, bmi) %>%
  mutate(cacs_log = log2(as.numeric(cacs)+1), .keep = "unused") %>%
  rename("Age" = "age",
         "Log2CACS" = "cacs_log",
         "BMI" = "bmi",
         "Sex" = "gender")

```

## Check

```{r}
bioheart_removed_variation %>% 
  rownames_to_column("record_id") %>%
  pivot_longer(cols = -record_id, names_to = "species", values_to = "value") %>% 
  mutate(class = plyr::mapvalues(species, 
                                 from = class_mapping$analyte, 
                                 to = class_mapping$Metabolite_Class, 
                                 warn_missing = F)) %>%
  group_by(class, record_id) %>%
  summarise(value = sum(value)) %>%
  mutate(record_id = as.numeric(record_id)) %>%
  left_join(key_clinical_vars, by = "record_id") %>%
  pivot_longer(cols = -c(class, record_id, value), names_to = "variable", values_to = "variable_value") %>%
  ggplot(aes(x = variable_value, y = value, colour = class)) + 
    geom_point(size = 0.5) + 
    theme_bw() +  
    geom_smooth(method = "lm", se = FALSE) +
    facet_wrap(~variable, scales = "free") +
    labs(title = "Removed variation with clinical variables", y = "Removed Variation", x = "") +
    guides(colour = "none")
```

```{r}
cholesterol <- readxl::read_excel("data/bioheart/bioheart_cholesterol_levels.xlsx")

key_cholesterol_vars <- cholesterol %>% select(record_id, tc, ldl, hdl, tgl) %>%
  rename("Total Cholesterol" = "tc",
         "LDL" = "ldl",
         "HDL" = "hdl",
         "TGL" = "tgl")

bioheart_removed_variation %>% 
  rownames_to_column("record_id") %>%
  pivot_longer(cols = -record_id, names_to = "species", values_to = "value") %>% 
  mutate(class = plyr::mapvalues(species, 
                                 from = class_mapping$analyte, 
                                 to = class_mapping$Metabolite_Class, 
                                 warn_missing = F)) %>%
  group_by(class, record_id) %>%
  summarise(value = sum(value)) %>%
  mutate(record_id = as.numeric(record_id)) %>%
  left_join(key_cholesterol_vars, by = "record_id") %>%
  pivot_longer(cols = -c(class, record_id, value), names_to = "variable", values_to = "variable_value") %>%
  ggplot(aes(x = variable_value, y = value, colour = class)) + 
    geom_point(size = 0.5) + 
    theme_bw() +  
    geom_smooth(method = "lm", se = FALSE) +
    facet_wrap(~variable, scales = "free") +
    labs(title = "Removed variation with cholesterol variables", y = "Removed Variation", x = "") +
    guides(colour = "none")
```

# Some extra files

```{r}
lipid_sds <- apply(log2(bioheart_BIO), 2, sd)
lipid_sds_df <- data.frame(lipid = names(lipid_sds),
                           sd = lipid_sds)
write.csv(lipid_sds_df, "lipid_sds.csv", row.names = FALSE)
```

